### 2020-06-17 report/sections/server/codegen.tex
Report: review chapter 3

\chapter{\fancyname{SessionTS}: Session Type API Generation for TypeScript}
\label{chap:codegen}

In this chapter, we present
\fancyname{SessionTS},
a toolchain for generating TypeScript APIs that developers can
use to write web applications that conform to their specified
communication protocol.
The toolchain is publicly available at \cite{repo}.

\fancyname{SessionTS} supports communication protocols that define 
\emph{server-centric} topologies, meaning that: 
\textbf{(1)} there is exactly one participant executed on the 
Node.js runtime; 
\textbf{(2)} all other participants run on the web browser; and
\textbf{(3)} all non-server participants only communicate with the server.
Examples of such protocols include the \tprotocol{Noughts and Crosses}
multiplayer game (\cref{section:evalgame}) and the \tprotocol{ATM} protocol.
We relax this assumption in \cref{part:general}.

>>>>

\chapter{\codegen: Session Type API Generation for TypeScript}
\label{chap:codegen}

In this chapter, we present
\codegen,
a toolchain for generating TypeScript APIs that developers can
use to write web applications that conform to their specified
communication protocol.
The toolchain is publicly available at \cite{repo}.

\codegen supports communication protocols that define 
\emph{server-centric} topologies, meaning that: 
\textbf{(1)} there is exactly one participant executed on the 
Node.js runtime; 
\textbf{(2)} all other participants run on the web browser; and
\textbf{(3)} all non-server participants only communicate with the server.
An example would be the \tprotocol{Noughts and Crosses}
multiplayer game (\cref{lst:game}).
We relax this assumption in \cref{part:general}.

### 2020-06-17 report/sections/server/codegen/implementation.tex
Report: review chapter 3

\fancyname{SessionTS} is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
Its rich standard library also simplifies many tasks: we use
the \texttt{argparse} package to generate an informative
command line interface (CLI) for developers to 
supply the correct information (\cref{lst:cmdline})
to use \fancyname{SessionTS},
and the \texttt{subprocess} package to 
invoke external tools, such as Scribble and the TypeScript Compiler.

@@

optional arguments:
  -h, --help     show this help message and exit
  -s SERVER, --server SERVER
                 Server role (only applicable for browser targets)
  -o OUTPUT, --output OUTPUT
                 Output directory for generation
\end{lstlisting}
\captionof{lstlisting}{Entry Point for \fancyname{SessionTS}}
\label{lst:cmdline}
\end{figure}

>>>>

\codegen is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
Its rich standard library also simplifies many tasks: we use
the \texttt{argparse} package to generate an informative
command line interface (CLI) for developers to 
supply the correct information (\cref{lst:cmdline})
to use \codegen
and the \texttt{subprocess} package to 
invoke external tools, such as Scribble and the TypeScript Compiler.

@@

optional arguments:
  -h, --help     show this help message and exit
  -s SERVER, --server SERVER
                 Server role (only applicable for browser targets)
  -o OUTPUT, --output OUTPUT
                 Output directory for generation
\end{lstlisting}
\captionof{lstlisting}{Entry Point for \codegen}
\label{lst:cmdline}
\end{figure}

### 2020-06-17 report/sections/server/codegen/workflow.tex
Report: review chapter 3

We motivate our development workflow from previous work 
\cite{PureScript2019, Hybrid2016} by extending the Scribble toolchain
and generating APIs that integrate the developer's 
application logic
into the execution of the communication automata.

@@

\item The developer supplies the communication protocol written in
Scribble (\cref{subsection:scribble}), 
stating the role (hereafter \textit{endpoint})
to generate APIs for,
and the code generation \textit{target}
(i.e. whether the role runs on the server or the web browser).

\item \fancyname{SessionTS} delegates to the 
Scribble toolchain for verifying the well-formedness of
the protocol and expects to receive a DOT graph representation of
the endpoint FSM (\cref{subsection:efsm}). 
\fancyname{SessionTS} parses the endpoint's 
interactions from the DOT graph and generates TypeScript APIs
for the developer (\cref{subsection:apigen}) 
tailored to the specified target.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{DevelopmentWorkflow}
\captionof{figure}{Overview of \fancyname{SessionTS} Development Workflow}
\label{fig:devworkflow}
\end{figure}

@@

We use the Scribble protocol description language, 
as presented in
\cite{Scribble}, for formalising the communication structure. This is
inspired by existing work on implementing session type theory 
in mainstream programming languages
\cite{Hybrid2016, PureScript2019, Python2017}. 
We use the variant of the Scribble language 
previously introduced in \cref{subsection:bgscribble}.

\subparagraph{Type declaration statements}
Specific to our TypeScript API generation toolchain,
the developer is \textit{not} required to explicitly
add type declaration statements for built-in types.
\cref{lst:adder} is a \textit{syntactically correct}
Scribble protocol as far as 
\fancyname{SessionTS} is concerned. 
Internally, \fancyname{SessionTS} inspects the protocol file
and parses existing type declarations using regular expressions
(or \textit{regex}) -- this is necessary to extract any
custom data types that will appear in the communication (for example,
\cref{lst:game}), and allows \fancyname{SessionTS} to inject
``boilerplate'' type declarations for built-in TypeScript types before
calling Scribble.

@@

Scribble expresses the EFSM using the 
DOT graph description language \cite{dot}, with each
communication action encoded as the label of the corresponding
state transition. 
\fancyname{SessionTS} uses the pydot library \cite{pydot}
to parse the graph into an internal representation of the EFSM.
We define an \texttt{EfsmBuilder} class with APIs designed for
constructing the EFSM representation by iterating over the 
state transitions from the DOT representation.

@@

For \fancyname{SessionTS}, we leverage the Jinja \cite{jinja} 
template engine library for code generation. 
We first construct templates for the TypeScript files we wish to generate,
specifying placeholders for dynamic content (to be extracted
from the \texttt{Endpoint} object); 
we then provide Jinja with the template path and the 
\texttt{Endpoint} object, and the template engine renders the
TypeScript code by filling in the dynamic placeholders. 
We show an example in \cref{lst:jinja}.

>>>>

We motivate our development workflow from previous work 
\cite{PureScript2019, Hybrid2016, LinearDecomp} 
by extending the Scribble toolchain
and generating APIs that integrate the developer's 
application logic
into the execution of the communication automata.

@@

\item The developer supplies the communication protocol written in
Scribble (\cref{subsection:scribble}), 
stating the role (hereafter \textit{endpoint})
to generate APIs for,
and the code generation \textit{target}
(i.e. whether the role runs on the back-end web server or 
the front-end web browser).

\item \codegen delegates to the 
Scribble toolchain for verifying the well-formedness of
the protocol and expects to receive a DOT graph representation of
the endpoint FSM (\cref{subsection:efsm}). 
\codegen parses the endpoint's 
interactions from the DOT graph and generates TypeScript APIs
for the developer (\cref{subsection:apigen}) 
tailored to the specified target.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{DevelopmentWorkflow}
\captionof{figure}{Overview of \codegen Development Workflow}
\label{fig:devworkflow}
\end{figure}

@@

We use the Scribble protocol description language, 
as presented in
\cite{Scribble}, for formalising the communication structure. This is
inspired by existing work on implementing session type theory 
in mainstream programming languages
\cite{Hybrid2016, PureScript2019, Python2017, LinearDecomp}. 
We use the variant of the Scribble language 
discussed in \cref{subsection:bgscribble}.

\subparagraph{Type declaration statements}
Specific to \codegen,
the developer is \textit{not} required to explicitly
add type declaration statements for built-in types.
\cref{lst:adder} is a \textit{syntactically correct}
Scribble protocol as far as 
\codegen is concerned. 
Internally, \codegen inspects the protocol file
and parses existing type declarations using regular expressions
(or \textit{regex}) -- this is necessary to extract any
custom data types that will appear in the communication (for example,
\cref{lst:game}), and allows \codegen to inject
``boilerplate'' type declarations for built-in TypeScript types before
calling Scribble.

@@

Scribble expresses the EFSM using the 
\textit{DOT} graph description language \cite{dot}, 
with each
communication action encoded as the label of the corresponding
state transition. 
\codegen uses the \textit{pydot} library \cite{pydot}
to parse the graph into an internal representation of the EFSM.
We define an \texttt{EfsmBuilder} class with APIs designed for
constructing the EFSM representation by iterating over the 
state transitions from the DOT representation.

@@

For \codegen, we leverage the \textit{Jinja} \cite{jinja} 
template engine library for code generation. 
We first construct templates for the TypeScript files 
we wish to generate,
specifying placeholders for dynamic content (to be extracted
from the \texttt{Endpoint} object); 
we then provide Jinja with the template path and the 
\texttt{Endpoint} object, and the template engine renders the
TypeScript code by filling in the dynamic placeholders. 
We show an example in \cref{lst:jinja}.
